Embedding Intuitionistic into Classical Logic

### 2022-10-14 FoSSaCS/fossacs-EIICL.tex
pass over intro + experiments finished

With this work we want to leverage the progress in classical reasoning for intuitionistic logic.
To this end, we propose the following methodology:
\begin{itemize}
	\item Translate an input formula $\varphi$ to a formula $\varphi^\#$, which is classically valid if and only if $\varphi$ is intuitionistically valid.
	\item Determine the classical validity of $\varphi^\#$ using a prover for classical logic, resulting in a proof or counter-model.
	\item Translate the classical proof/counter-model of $\varphi^\#$ to a corresponding intuitionistic proof/counter-model of $\varphi$.
\end{itemize}
\textbf{Contributions.} We now overview the main results of this paper (which we discuss in more detail in Sections~\ref{sec:overview} and~\ref
\begin{itemize}
	\item We establish a small model property for counter-models to validity in intuitionistic propositional logic.
	\item Based on the small model property, we provide an embedding of intuitionistic propositional logic into classical propositional logic and into quantified boolean formulas, i.e., we provide an 
	\item We discuss the reduction of counter-models in intuitionistic first-order logic and, based on this reduction, present an embedding intuitionistic into classical first-order logic.
	\item We provide an experimental evaluation on the use of provers for classical first-order logic for determining intuitionistic validity.
 We show that our approach can compete with state-of-the-art provers for certain classes of benchmarks, in particular when the intuitionistic content is low.
\end{itemize}

@@

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a formula $\varphi^\#$ with $\mathcal |\varphi^\#|\in\mathcal O(|\mathcal R|\cdot\exp(|\mathcal X|\log|\mathcal X|))$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid.
\end{lemma}

@@

A detailed account of the translation can be found in definition~\ref{def:fo-translation}. Finally, we benchmark our translation by translating all problems in the ILTP problem set~\cite{iltp} and applying the Vampire theorem prover~\cite{kovacs2013first} to the translated instances. While this method falls short of the absolute state-of-the-art in general, for certain subsets of benchmarks it beats existing provers, in particular when the intuitionistic content is low. Since there is still a lot of room for improvement this is a hopeful first sign.

@@

\section{Benchmark}

@@

Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results can be seen in Figure \ref{fig:a}. In general our approach falls short of the state of the art. However when binning for $|\mathcal Q|$ and $|\mathcal X|$ as seen in Figure~\ref{fig:b} we observe good performance for $|\mathcal Q| = 0$ and $|\mathcal X| \leq 3$. Figure~\ref{fig:c} shows that we solve 52 problems not solved by the existing provers.
\begin{figure}
	\centering
	\begin{tabular}{c|c|lr|lr|lr|lr|c}
		$|\mathcal Q|$&$|\mathcal X|$&\multicolumn{2}{c}{Embedding}&\multicolumn{2}{c}{Embedding (idem.)}&\multicolumn{2}{c}{ileanCoP}&\multicolumn{2}{c}{nanoCoP-i 2.0}&Total\\\hline
		0&0		&157&88.2\%	&157&88.2\%		&128&71.9\%		&61&38.1\%	&178\\
		0&1-2	&119&29.8\%&130&32.5\%		&108&27.0\%		&96&24.0\%	&400\\
		0&3-5	&16&15.2\%	&17&16.2\%		&50&47.6\%		&40&28.1\%	&105\\
		0&6+	&27&38.6\%	&11&15.7\%		&20&28.6\%		&11&15.7\%	&70\\
		1-2&0-2	&84&29.3\%	&81&28.2\%		&227&79.1\%		&221&77.0\%	&287\\
		1-2&3-5	&80&25.6\%	&77&24.7\%		&141&45.2\%		&122&39.1\%	&312\\
		1-2&6+	&22&21.8\%	&21&20.8\%		&39&38.6\%		&37&36.6\%	&101\\
		3-5&0-5	&38&17.0\%	&40&17.9\%		&125&55.8\%		&110&49.1\%	&224\\
		3-5&6+  &20&8.4\%	&18&7.6\%		&64&27.0\%		&86&36.3\%	&237\\
		6+&0+	&8&1.1\%	&5&0.7\%		&73&9.7\%		&74&9.8\%	&755
	\end{tabular}
	\caption{Solved problems binned by $|\mathcal Q|$ and $|\mathcal X|$.}
	\label{fig:b}
\end{figure}

 Therefore our tool is already useful if it is known that $|\mathcal X|$ and $|\mathcal Q|$ are small. We see that idempotence does not improve performance. On the other hand from a model-theoretic perspective it is quite powerful. We suspect that a more clever encoding of this property could improve our results. There are other restrictions on models we could consider, based on structural properties of the formula, e.g. the relation between occurrences of different atoms.

>>>>

With this work we want to leverage the progress in classical reasoning for intuitionistic logic.
To this end, we study embeddings of intuitionistic into classical logic and propose the following methodology:
\begin{itemize}
	\item Translate an input formula $\varphi$ to a formula $\varphi^\#$, which is classically valid if and only if $\varphi$ is intuitionistically valid.
	\item Determine the classical validity of $\varphi^\#$ using a prover for classical logic, resulting in a proof or counter-model.
	\item Translate the classical proof/counter-model of $\varphi^\#$ to a corresponding intuitionistic proof/counter-model of $\varphi$.
\end{itemize}
\subsection*{Contributions} 
We now overview the main results of this paper (which we discuss in more detail in Sections~\ref{sec:overview} and~\ref{sec:evaluation}):
\begin{itemize}
	\item We establish a small model property for counter-models to validity in intuitionistic propositional logic.
	\item Based on the small model property, we provide an embedding of intuitionistic propositional logic into classical propositional logic and into quantified boolean formulas.
	\item We discuss the reduction of counter-models in intuitionistic first-order logic and, based on this reduction, present an embedding of intuitionistic into classical first-order logic.
	\item We provide an experimental evaluation on the use of first-order provers for classical logic for determining intuitionistic validity.
 We show that our approach can compete with state-of-the-art provers for certain classes of benchmarks, in particular when the intuitionistic content is low.
\end{itemize}

@@

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a formula $\varphi^\#$ with $\mathcal |\varphi^\#|\in\mathcal O(|\mathcal R|\cdot 2^{|\mathcal X|\log|\mathcal X|})$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid.
\end{lemma}

@@

A detailed account of the translation can be found in Definition~\ref{def:fo-translation}. Finally, we benchmark our translation by translating all problems in the ILTP problem set~\cite{iltp} and applying the Vampire theorem prover~\cite{kovacs2013first} to the translated instances. While this method falls short of the absolute state-of-the-art in general, for certain subsets of benchmarks it beats existing provers, in particular when the intuitionistic content is low. Since there is still a lot of room for improvement this is a hopeful first sign.

@@

\section{Experimental Evaluation}
\label{sec:evaluation}

@@

Two versions of the translation were tested, one with and one without the additional axioms encoding idempotence. The results are stated in Figure \ref{fig:a}. 
In general, our approach falls short of the state of the art. 
However when grouping the benchmarks by $|\mathcal Q|$ and $|\mathcal X|$, see Figure~\ref{fig:b},
we observe good performance of our approach for $|\mathcal Q| = 0$ and $|\mathcal X| \leq 2$ resp. $|\mathcal X| \geq 6$.
\todo{Note that I added $\ge 6$.}
Figure~\ref{fig:c} shows that our approach solved 52 problems not solved by the state-of-the-art provers.
\begin{figure}
	\centering
	\begin{tabular}{c|c|lr|lr|lr|lr|c}
		$|\mathcal Q|$&$|\mathcal X|$&\multicolumn{2}{c}{Embedding}&\multicolumn{2}{c}{Embedding (idem.)}&\multicolumn{2}{c}{ileanCoP}&\multicolumn{2}{c}{nanoCoP-i 2.0}&Total\\\hline
		0&0		&157&88.2\%	&157&88.2\%		&128&71.9\%		&61&38.1\%	&178\\
		0&1-2	&119&29.8\%&130&32.5\%		&108&27.0\%		&96&24.0\%	&400\\
		0&3-5	&16&15.2\%	&17&16.2\%		&50&47.6\%		&40&28.1\%	&105\\
		0&6+	&27&38.6\%	&11&15.7\%		&20&28.6\%		&11&15.7\%	&70\\
		1-2&0-2	&84&29.3\%	&81&28.2\%		&227&79.1\%		&221&77.0\%	&287\\
		1-2&3-5	&80&25.6\%	&77&24.7\%		&141&45.2\%		&122&39.1\%	&312\\
		1-2&6+	&22&21.8\%	&21&20.8\%		&39&38.6\%		&37&36.6\%	&101\\
		3-5&0-5	&38&17.0\%	&40&17.9\%		&125&55.8\%		&110&49.1\%	&224\\
		3-5&6+  &20&8.4\%	&18&7.6\%		&64&27.0\%		&86&36.3\%	&237\\
		6+&0+	&8&1.1\%	&5&0.7\%		&73&9.7\%		&74&9.8\%	&755
	\end{tabular}
	\caption{Solved problems grouped by $|\mathcal Q|$ and $|\mathcal X|$.}
	\label{fig:b}
\end{figure}

We conclude that our approach already has merits when $|\mathcal X|$ and $|\mathcal Q|$ are small. 
We see that encoding idempotence does help to solve some additional benchmarks, but in general does not improve performance.
However, idempotence is a quite powerful concept from a model-theoretic perspective. We suspect that a more clever encoding of this property could improve our results. There are other restrictions on models we could consider, based on structural properties of the formula, e.g. the relation between occurrences of different atoms.
